Harnessing generative AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-06-10 docs/_tex/manuscript.tex
KBM manuscript edits

There are thousands of human phenotypes which are linked to genetic
variation. These range from the benign (white eyelashes) to the deadly
(respiratory failure). The Human Phenotype Ontology has categorised all
human phenotypic variation into an unified framework that defines the
relationships between them (e.g.~missing arms and missing legs are both
abnormalities of the limb). This has made it possible to perform
phenome-wide analyses, e.g.~to prioritise which make the best candidates
for gene therapies. However, there is currently limited metadata
describing the effects / severity of these phenotypes. With
\textgreater17500 phenotypic abnormalities across \textgreater8600 rare
diseases, manual curation of such phenotypic annotations by experts
would be exceedingly labour-intensive and time-consuming. Leveraging
advances in artificial intelligence, we employed the OpenAI GPT-4 large
language model (LLM) to systematically annotate the severity of all
phenotypic abnormalities in the HPO. First, we benchmarked the
generative LLM annotations against ground-truth labels within the HPO
(e.g.~phenotypes in the `Cancer' HPO branch were annotating as causing
cancer by GPT-4). True positive recall rates across different
annotations ranged from 89-100\% (mean=96\%), clearly demonstrating the
ability of GPT-4 to automate the curation process with a high degree of
fidelity. Using a novel approach, we developed a severity scoring system
that incorporates both the nature of the phenotype outcome and the
frequency of its occurrence. These severity metrics will enable efforts
to systematically prioritise which human phenotypes are most detrimental
to human health, and best targets for therapeutic intervention.

@@

To evaluate the consistency of responses generated by GPT-4 793
phenotypes were annotated multiple times. For a subset of phenotypes
with known expected annotations, true positive rates were calculated to
assess recall. Additionally, based on the clinical characteristics and
their occurrence, we have quantified the severity of each phenotype,
providing an example of how these clinical characteristic annotations
can be used to guide prioritisation of gene therapy trials. Ultimately,
we hope that our resource will be of utility to those working in rare
diseases, as well as the wider healthcare community.

@@

\caption{\label{fig-occurrence}GPT-4 was able to annotate all human
phenotypes based on whether they are always/often/rarely/never
associated with different clinical characteristics. \textbf{a}, An
example of the prompt input given to to GPT-4. The phenotypes listed in
the second to last sentence (\emph{italicised}) were changed to allow
all HPO phenotypes to be annotated. \textbf{b}, Stacked bar plot showing
the proportion of the occurrence of each clinical characteristic across
all annotated HPO phenotypes. The terms shown on the x-axis are the
clinical characteristics for which GPT-4 was asked to determine whether
each phenotype caused them.}

@@

We employed the OpenAI GPT-4 model with Python to annotate 17502 terms
within the HPO (v2024-02-08) (Gargano et al., 2024; Köhler et al.,
2021). Our annotation framework was developed based on previously
defined criteria for classifying disease severity (Lazarin et al.,
2014). We sought to evaluate the impact of phenotypes on factors
including intellectual disability, death, impaired mobility, physical
malformations, blindness, sensory impairments, immunodeficiency, cancer,
reduced fertility, and congenital onset. Through prompt design we found
that the performance of GPT-4 improved when we incorporated a scale
associated with each effect and required a justification for each
response. For each effect, we asked about the frequency of its
occurrence - whether it never, rarely, often, or always occurred.
Framing the queries in this way served two purposes. First, this helped
to constrain the responses of GPT-4 to a specific range of values,
making answers more consistent and amenable to downstream data analysis.
Second, it served to overcome one of the main limitations noted by
Lazarin et al. (2014) as they did not collect information on how the
frequency of each disease affected their decision making when generating
severity annotations.

Phenotype outcome occurrence varied across annotation categories.
\textgreater50\% of phenotypes never caused blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility or intellectual
disability. Only a minority of phenotypes (21.7\%) never had a
congenital onset, which is expected as rare disorders tend to be early
onset genetic conditions Fig.~\ref{fig-occurrence}.

@@

\caption{\label{fig-checks}GPT-4 annotations are consistent and accurate
across annotations. \textbf{a}, Barplot showing the annotation
consistency within phenotypes that were annotated more than once. In the
lenient metric, annotations were collapsed into two groups
(`always'/`often' and `never'/`rarely'). For a given clinical
characteristic within a given phenotype, if an annotation was always
within the same group it was considered consistent. In the stringent
metric, all four annotation categories were considered to be different
from one another. Thus, annotations were only defined as consistent if
they were all identical. The blue dashed line indicates the probability
of two annotations being consistent by chance in the lenient metric
(\textasciitilde1/2). The gold dashed line is the probability of two
annotations being consistent by chance in the stringent metric
(\textasciitilde1/4). \textbf{b}, Bar plot of the true positive rate for
each annotation. The labels above each bar indicate the number of
phenotypes tested.}

@@

While individual annotations are informative, we wanted to be able to
distil the severity of each phenotype into a single score. Quantifying
the overall severity of phenotypes can have important implications for
diagnosis, prognosis, and treatment. It may also guide the
prioritisation of gene therapy trials for phenotypes with the most
severe outcomes and thus the most urgent need. Importantly, the values
reflected the severity of each outcome based on both the type of outcome
itself and its frequency within a particular phenotype. For instance, a
phenotype always causing death would have a higher multiplied value than
a phenotype often causing reduced fertility (see
Table~\ref{tbl-metric-weights}). First, we created a dictionary to map
each phenotype outcome (e.g.~blindness) and its frequency (always,
often, rarely, never) to numeric values from 0-3. Then, the phenotype
outcome values were multiplied by phenotype outcome weights. Next, we
computed an average score for each phenotype by aggregating the
multiplied values across all phenotype outcomes and then calculating the
mean. This was then normalised by the theoretical maximum severity
score, so that all phenotypes were on a 0-100 severity scale (where 100
is the most severe phenotype possible). This average normalised score
represents the overall severity of the phenotype based on the severity
of its individual outcomes.

Based on these scores we evaluated the top 50 severe phenotypes. One of
the most severe phenotype was `Anencephaly' (HP:0002323) with a
composite severity score of 45. Anencephaly is a birth defect where the
baby is born without a portion of its brain and skull, often these
babies are stillborn. In fact, many of the most severe phenotypes were
related to developmental brain and neural tube defects. Comparison of
the severity scores for each response, across the phenotype outcomes
annotated, revealed consistent trends: as the response of the phenotype
outcome increased (from never to always), the severity score also
increased (Supplementary Fig.~\ref{fig-severity-boxplot}). We also
evaluated the severity score distribution by HPO branch and calculated
the mean severity score using all phenotypes within each major HPO
branch (Fig.~\ref{fig-severity-histo}). The HPO branch with the greatest
mean severity score was `Abnormal cellular phenotype' (mean=17),
followed by `Neoplasm' (mean=15.2), which would include the highly
ranked phenotypes seen in Figure~\ref{fig-top-phenos}.

@@

\subsubsection{Correlations between severity
metrics}\label{correlations-between-severity-metrics}

We found that some severity metrics were correlated with one another,
with a mean Pearson correlation of 0.2 across all individual metrics
(see Figure~\ref{fig-metric-corplot}). In particular, blindness and
sensory impairment were highly correlated with one another (r=0.62,
p=0). Some metrics drove the composite severity score more than other,
which is a reflection of both our per-metric weighting scheme, response
type frequencies, and the correlation structure between metrics.
Overall, impaired mobility seemed to be the strongest driver of the
composite severity score with a Pearson correlation of 0.6001824,
followed by intellectual disability (r=0.59) and death (r=0.56).

@@

A key contribution of our study is the introduction of a quantitative
severity scoring system that integrates both the nature of the phenotype
outcome and the frequency of its occurrence. By encoding the concept of
severity in this way, we are able to prioritise phenotypes based on
their impact on patients. The methodology allowed us to transition from
low-throughput qualitative assessments of severity (e.g. Lazarin et al.
(2014)) to high-throughput quantitative assessments of severity. One of
the most severe phenotypes in the HPO is `Fetal akinesia sequence' (FAS;
HP:0001989, severity score= 43.9), and extremely rare condition that is
almost always lethal. FAS is a complex, multi-system phenotype that can
be caused by at least 24 different genetic disorders. Despite the
complex and heterogeneous aetiology of this phenotype, GPT-4 was able to
provide accurate annotations alongside explainable justifications for
those annotations (see Table~\ref{tbl-fas}). For example, this phenotype
almost always results in death, either \emph{in utero} or shortly after
birth. Not only did GPT-4 correctly provide the annotation death as
`always', when asked whether FAS causes sensory impairments it provided
the response `always' with the justification `Fetal akinesia sequence
typically results in severe sensory impairment due to neurodevelopmental
disruption.' Neurodevelopmental disruption is indeed a hallmark
component of FAS (e.g.~hydrocephalus, cerebellar hypoplasia) that causes
severe impairments across multiple sensory systems (Chen, 2012). This
demonstrates that GPT-4 was able to recover the correct chain of
causality from phenotype to clinical outcome.

@@

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each effect and required a justification for each response. We
asked how frequently the given phenotype directly causes each clinical
characteristic - whether it never, rarely, often, or always occurred.
This design helps to constrain the potential responses of GPT-4 and thus
make it more amenable to machine-readable post-processing. It also
serves to address one of its key limitations from the Lazarin et al.
(2014) survey, namely the lack information on how clinical
characteristic frequency affected the clinicians' severity annotations.
Here, we can instead use the frequency values to generate more precise
annotations and downstream severity ranking scores.

@@

\caption{\label{tbl-true-positives}For each clinical characteristic, any
terms that matched}

@@

The GPT-4 generated annotation occurrences were converted into a
semi-quantitative scoring system, with `always' corresponding to 3,
`often' to 2, `rarely' to 1, and `never' to 0. These scores were then
weighted by a severity metric on a scale of 1-5, with 5 representing the
highest severity, as determined by the provided annotations
(Table~\ref{tbl-metric-weights}). Subsequently, the weighted scores
underwent normalisation to yield a final quantitative severity score
ranging from 0-100, with 100 signifying the maximum severity score
attainable.

@@

@@

The decision tree algorithm used in Lazarin et al. (2014) was adapted
here for use with the GPT-4 annotations. This algorithm first assigned
each annotation metric to a tier, where Tier 1 indicated the most severe
annotation metrics and Tier 4 indicated the least severe annotation
metrics (`death'=1, `intellectual disability'=1, `impaired mobility'=2,
`physical malformations'=2, `blindness'=3, `sensory impairments'=3,
`immunodeficiency'=3, `cancer'=3, `reduced fertility'=4). If a phenotype
often or always caused more than one Tier 1 annotation, it was assigned
a severity class of ``Profound''. If the phenotype often or always
caused only one Tier 1 annotation, it was assigned a severity class of
``Severe''. A ``Severe'' class assignment was also assigned if the
phenotype often or always caused three or more Tier 2 and Tier3
annotations. If the phenotype often or always caused at least one Tier 2
annotation, it was assigned a severity class of ``Moderate''. All
remaining phenotypes were was assigned a severity class of ``Mild''. In
cases where the phenotype mapped to more than one class, only the most
severe class was used. This procedure is implemented within the function
\texttt{HPOExplorer::gpt\_annot\_class}.

\subsubsection{Correlations between severity
metrics}\label{correlations-between-severity-metrics-1}

To assess the correlation structure between each severity metric, as
well as between the composite severity score and each metric, we
computed Pearson correlation coefficients for all pairwise combinations
of these variables using the numerically encoded metric values. The
correlation matrix was visualised using a heatmap, with the colour
intensity representing the strength of the correlation
(Figure~\ref{fig-metric-corplot}).

@@

The GPT-4 annotations for all HPO phenotypes are made available through
the R function \texttt{HPOExplorer::gpt\_annot\_read} or in CSV format
at:

@@

\caption{\label{fig-consist-vs-ontLvl}Relationship between the
consistency of GPT-4 annotations (using the stringent criterion) and the
level of each phenotype within the HPO ontology (with the number of
phenotypes in parentheses). Greater ontology levels (x-axis) indicate
more specific phenotypes. The subtitle indicates summary statistics for
the overall relationship between HPO level and the proportion of
phenotypes that were annotated consistently. The p-values above each bar
indicate whether the distribution of consistent/inconsistent
annotations, within a given HPO level, significantly deviate from the
expected null distribution.}

@@

\caption{\label{fig-severity-boxplot}Boxplot showing the relationship
between composite severity score (y-axis) and the frequency response
categories within each annotation type.}

@@

\caption{\label{fig-metric-corplot}Pearson correlations between each
individual severity metric and the composite severity score
(`severity\_score\_gpt').}

@@

\caption{\label{tbl-annotations}Table of GPT-4 annotations for all Human
Phenotype Ontology (HPO) phenotypes in Figure~\ref{fig-top-phenos}. For
each phenotype, this includes the name of the phenotype (`hpo\_name'),
the ID of the phenotype (`hpo\_id'), the frequency of each annotation
(always, often, rarely, never), and the justification for each
annotation ('\ldots\_justification'). These results can also be
downloaded programmatically using the R function
\texttt{HPOExplorer::gpt\_annot\_check}.}

@@

\caption{\label{tbl-fas}Severity nnotations generated for GPT-4
annotations for the HPO phenotype `Fetal akinesia sequence'
(HP:000198).}

>>>>

There are thousands of human phenotypes which are linked to genetic
variation. These range from the benign (white eyelashes) to the deadly
(respiratory failure). The Human Phenotype Ontology has categorised all
human phenotypic variation into an unified framework that defines the
relationships between them (e.g.~missing arms and missing legs are both
abnormalities of the limb). This has made it possible to perform
phenome-wide analyses, e.g.~to prioritise which make the best candidates
for gene therapies. However, there is currently limited metadata
describing the clinical characteristics / severity of these phenotypes.
With \textgreater17500 phenotypic abnormalities across \textgreater8600
rare diseases, manual curation of such phenotypic annotations by experts
would be exceedingly labour-intensive and time-consuming. Leveraging
advances in artificial intelligence, we employed the OpenAI GPT-4 large
language model (LLM) to systematically annotate the severity of all
phenotypic abnormalities in the HPO. Phenotypic severity was defined
using a set of clinical characteristics and their frequency of
occurrence. First, we benchmarked the generative LLM clinical
characteristic annotations against ground-truth labels within the HPO
(e.g.~phenotypes in the `Cancer' HPO branch were annotating as causing
cancer by GPT-4). True positive recall rates across different clinical
characteristics ranged from 89-100\% (mean=96\%), clearly demonstrating
the ability of GPT-4 to automate the curation process with a high degree
of fidelity. Using a novel approach, we developed a severity scoring
system that incorporates both the nature of the clinical characteristic
and the frequency of its occurrence. These clinical characteristic
severity metrics will enable efforts to systematically prioritise which
human phenotypes are most detrimental to human health, and best targets
for therapeutic intervention.

@@

To evaluate the consistency of responses generated by GPT-4 793
phenotypes were annotated multiple times. For a subset of phenotypes
with known expected clinical characteristics, true positive rates were
calculated to assess recall. Additionally, based on the clinical
characteristics and their occurrence, we have quantified the severity of
each phenotype, providing an example of how these clinical
characteristic annotations can be used to guide prioritisation of gene
therapy trials. Ultimately, we hope that our resource will be of utility
to those working in rare diseases, as well as the wider healthcare
community.

@@

\caption{\label{fig-occurrence}GPT-4 was able to annotate all human
phenotypes based on whether they are always/often/rarely/never
associated with different clinical characteristics. \textbf{a} An
example of the prompt input given to to GPT-4. The phenotypes listed in
the second to last sentence (\emph{italicised}) were changed to allow
all HPO phenotypes to be annotated. \textbf{b} Stacked bar plot showing
the proportion of the occurrence of each clinical characteristic across
all annotated HPO phenotypes. The terms shown on the x-axis are the
clinical characteristics for which GPT-4 was asked to determine whether
each phenotype caused them.}

@@

We employed the OpenAI GPT-4 model with Python to annotate 17502 terms
within the HPO (v2024-02-08) (Gargano et al., 2024; Köhler et al.,
2021). Our annotation framework was developed based on previously
defined criteria for classifying disease severity (Lazarin et al.,
2014). We sought to evaluate the impact of phenotypes on factors
including intellectual disability, death, impaired mobility, physical
malformations, blindness, sensory impairments, immunodeficiency, cancer,
reduced fertility, and congenital onset. Through prompt design we found
that the performance of GPT-4 improved when we incorporated a scale
associated with each clinical characteristic and required a
justification for each response. For each clinical characteristic, we
asked about the frequency of its occurrence - whether it never, rarely,
often, or always occurred. Framing the queries in this way served two
purposes. First, this helped to constrain the responses of GPT-4 to a
specific range of values, making answers more consistent and amenable to
downstream data analysis. Second, it served to overcome one of the main
limitations noted by Lazarin et al. (2014) as they did not collect
information on how the frequency of each disease affected their decision
making when generating severity annotations.

Clinical characteristic occurrence varied across annotation categories.
\textgreater50\% of phenotypes never caused blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility or intellectual
disability. Only a minority of phenotypes (21.7\%) never had a
congenital onset, which is expected as rare disorders tend to be early
onset genetic conditions (Fig.~\ref{fig-occurrence}).

@@

\caption{\label{fig-checks}GPT-4 annotations are consistent and accurate
across annotations. \textbf{a} Barplot showing the annotation
consistency within phenotypes that were annotated more than once. In the
lenient metric, annotations were collapsed into two groups
(`always'/`often' and `never'/`rarely'). For a given clinical
characteristic within a given phenotype, if an annotation was always
within the same group it was considered consistent. In the stringent
metric, all four annotation categories were considered to be different
from one another. Thus, annotations were only defined as consistent if
they were all identical. The blue dashed line indicates the probability
of two annotations being consistent by chance in the lenient metric
(\textasciitilde1/2). The gold dashed line is the probability of two
annotations being consistent by chance in the stringent metric
(\textasciitilde1/4). \textbf{b} Bar plot of the true positive rate for
each annotation. The labels above each bar indicate the number of
phenotypes tested.}

@@

While individual annotations are informative, we wanted to be able to
distil the severity of each phenotype into a single score. Quantifying
the overall severity of phenotypes can have important implications for
diagnosis, prognosis, and treatment. It may also guide the
prioritisation of gene therapy trials for phenotypes with the most
severe clinical characteristics and thus the most urgent need.
Importantly, the values reflected the severity of each clinical
characteristic based on both the type of characteristic itself and its
frequency within a particular phenotype. For instance, a phenotype
always causing death would have a higher multiplied value than a
phenotype often causing reduced fertility (see
Table~\ref{tbl-metric-weights}). First, we created a dictionary to map
each clinical characteristic (e.g.~blindness) and its frequency (always,
often, rarely, never) to numeric values from 0-3. Then, the clinical
characteristic values were multiplied by weights. Next, we computed an
average score for each phenotype by aggregating the multiplied values
across all clinical characteristics and then calculating the mean. This
was then normalised by the theoretical maximum severity score, so that
all phenotypes were on a 0-100 severity scale (where 100 is the most
severe phenotype possible). This average normalised score represents the
overall severity of the phenotype based on the severity of its
individual clinical characteristics.

Based on these scores we evaluated the top 50 severe phenotypes. One of
the most severe phenotype was `Anencephaly' (HP:0002323) with a
composite severity score of 45. Anencephaly is a birth defect where the
baby is born without a portion of its brain and skull, often these
babies are stillborn. In fact, many of the most severe phenotypes were
related to developmental brain and neural tube defects. Comparison of
the severity scores for each response, across the clinical
characteristics annotated, revealed consistent trends: as the response
of the clinical characteristic increased (from never to always), the
severity score also increased (Supplementary
Fig.~\ref{fig-severity-boxplot}). We also evaluated the severity score
distribution by HPO branch and calculated the mean severity score using
all phenotypes within each major HPO branch
(Fig.~\ref{fig-severity-histo}). The HPO branch with the greatest mean
severity score was `Abnormal cellular phenotype' (mean=17), followed by
`Neoplasm' (mean=16.7), which would include the highly ranked phenotypes
seen in Figure~\ref{fig-top-phenos}.

@@

\subsubsection{Correlations between clinical characteristic severity
metrics}\label{correlations-between-clinical-characteristic-severity-metrics}

We found that some clinical characteristic severity metrics were
correlated with one another, with a mean Pearson correlation of 0.2
across all individual metrics (see Figure~\ref{fig-metric-corplot}). In
particular, blindness and sensory impairment were highly correlated with
one another (r=0.62, p=0). Some metrics drove the composite severity
score more than other, which is a reflection of both our per-metric
weighting scheme, response type frequencies, and the correlation
structure between metrics. Overall, impaired mobility seemed to be the
strongest driver of the composite severity score with a Pearson
correlation of 0.6001824, followed by intellectual disability (r=0.59)
and death (r=0.56).

@@

A key contribution of our study is the introduction of a quantitative
severity scoring system that integrates both the nature of the clinical
characteristic and the frequency of its occurrence. By encoding the
concept of severity in this way, we are able to prioritise phenotypes
based on their impact on patients. The methodology allowed us to
transition from low-throughput qualitative assessments of severity (e.g.
Lazarin et al. (2014)) to high-throughput quantitative assessments of
severity. One of the most severe phenotypes in the HPO is `Fetal
akinesia sequence' (FAS; HP:0001989, severity score= 43.9), and
extremely rare condition that is almost always lethal. FAS is a complex,
multi-system phenotype that can be caused by at least 24 different
genetic disorders. Despite the complex and heterogeneous aetiology of
this phenotype, GPT-4 was able to provide accurate annotations alongside
explainable justifications for those annotations (see
Table~\ref{tbl-fas}). For example, this phenotype almost always results
in death, either \emph{in utero} or shortly after birth. Not only did
GPT-4 correctly provide the annotation death as `always', when asked
whether FAS causes sensory impairments it provided the response `always'
with the justification `Fetal akinesia sequence typically results in
severe sensory impairment due to neurodevelopmental disruption.'
Neurodevelopmental disruption is indeed a hallmark component of FAS
(e.g.~hydrocephalus, cerebellar hypoplasia) that causes severe
impairments across multiple sensory systems (Chen, 2012). This
demonstrates that GPT-4 was able to recover the correct chain of
causality from phenotype to clinical characteristic.

@@

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each clinical characteristic and required a justification for each
response. We asked how frequently the given phenotype directly causes
each clinical characteristic - whether it never, rarely, often, or
always occurred. This design helps to constrain the potential responses
of GPT-4 and thus make it more amenable to machine-readable
post-processing. It also serves to address one of its key limitations
from the Lazarin et al. (2014) survey, namely the lack information on
how clinical characteristic frequency affected the clinicians' severity
annotations. Here, we can instead use the frequency values to generate
more precise annotations and downstream severity ranking scores.

@@

\caption{\label{tbl-true-positives}The HPO branches and their
descendants used as true positives for each clinical characteristic.}

@@

The GPT-4 generated clinical characteristic occurrences were converted
into a semi-quantitative scoring system, with `always' corresponding to
3, `often' to 2, `rarely' to 1, and `never' to 0. These scores were then
weighted by a severity metric on a scale of 1-5, with 5 representing the
highest severity, as determined by the provided clinical characteristics
(Table~\ref{tbl-metric-weights}). Subsequently, the weighted scores
underwent normalisation to yield a final quantitative severity score
ranging from 0-100, with 100 signifying the maximum severity score
attainable.

@@

\vspace{2em}

@@

The decision tree algorithm used in Lazarin et al. (2014) was adapted
here for use with the GPT-4 clinical characteristic annotations. This
algorithm first assigned each clinical chacteristic to a tier, where
Tier 1 indicated the most severe clinical characteristics and Tier 4
indicated the least severe clinical characteristics (`death'=1,
`intellectual disability'=1, `impaired mobility'=2, `physical
malformations'=2, `blindness'=3, `sensory impairments'=3,
`immunodeficiency'=3, `cancer'=3, `reduced fertility'=4). If a phenotype
often or always caused more than one Tier 1 clinical characteristic, it
was assigned a severity class of ``Profound''. If the phenotype often or
always caused only one Tier 1 clinical characteristic, it was assigned a
severity class of ``Severe''. A ``Severe'' class assignment was also
assigned if the phenotype often or always caused three or more Tier 2
and Tier3 clinical characteristics. If the phenotype often or always
caused at least one Tier 2 clinical characteristic, it was assigned a
severity class of ``Moderate''. All remaining phenotypes were was
assigned a severity class of ``Mild''. In cases where the phenotype
mapped to more than one class, only the most severe class was used. This
procedure is implemented within the function
\texttt{HPOExplorer::gpt\_annot\_class}.

\subsubsection{Correlations between clinical characteristic severity
metrics}\label{correlations-between-clinical-characteristic-severity-metrics-1}

To assess the correlation structure between each clinical characteristic
severity metric, as well as between the composite severity score and
each metric, we computed Pearson correlation coefficients for all
pairwise combinations of these variables using the numerically encoded
metric values. The correlation matrix was visualised using a heatmap,
with the colour intensity representing the strength of the correlation
(Figure~\ref{fig-metric-corplot}).

@@

The GPT-4 clinical characteristic annotations for all HPO phenotypes are
made available through the R function
\texttt{HPOExplorer::gpt\_annot\_read} or in CSV format at:

@@

\caption{\label{fig-consist-vs-ontLvl}Relationship between the
consistency of GPT-4 clinical characterstic annotations (using the
stringent criterion) and the level of each phenotype within the HPO
ontology (with the number of phenotypes in parentheses). Greater
ontology levels (x-axis) indicate more specific phenotypes. The subtitle
indicates summary statistics for the overall relationship between HPO
level and the proportion of phenotypes that were annotated consistently.
The p-values above each bar indicate whether the distribution of
consistent/inconsistent annotations, within a given HPO level,
significantly deviate from the expected null distribution.}

@@

\caption{\label{fig-severity-boxplot}Boxplot showing the relationship
between composite severity score (y-axis) and the frequency response
categories within each clinical characteristic type.}

@@

\caption{\label{fig-metric-corplot}Pearson correlations between each
individual clinical characteristic severity metric and the composite
severity score (`severity\_score\_gpt').}

@@

\caption{\label{tbl-annotations}Table of GPT-4 clinical characterstic
annotations for all Human Phenotype Ontology (HPO) phenotypes in
Figure~\ref{fig-top-phenos}. For each phenotype, this includes the name
of the phenotype (`hpo\_name'), the ID of the phenotype (`hpo\_id'), the
frequency of each annotation (always, often, rarely, never), and the
justification for each annotation ('\ldots\_justification'). These
results can also be downloaded programmatically using the R function
\texttt{HPOExplorer::gpt\_annot\_check}.}

@@

\caption{\label{tbl-fas}Severity nnotations generated for GPT-4 clinical
characteristic annotations for the HPO phenotype `Fetal akinesia
sequence' (HP:000198).}
